Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
a few notes

@@

When we need to store into our objects pointers to other objects, we
need to be careful: if we were to directly store regular addresses valid
in the local segment, then the objects' binary content would differ
between segments.  This would prevent any sharing.  Instead, we store
pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
of the segment.  A $SO$ can then be interpreted from any segment by
adding it to the base address of that segment.\footnote{An alternative
model would be to simulate threads using several independent processes
in which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the state
(file descriptors, external libraries, ...) is shared.} The result is
called a \emph{Linear Address (LA)}. This is illustrated in
figure~\ref{fig:Segment-Addressing}.

@@

Conflicts can be resolved by aborting one of the two transactions, or in
some cases by pausing one of them until the other commits.  If a commit
succeeds, we copy the changes to modified objects so that they become
visible in other segments.  Note that this only includes objects that
existed before the transaction and have been modified by it.  It does
not includes all the new objects created by the current transaction
(there is typically a very high turn-around rate in dynamic languages).
Our garbage collector allocates new objects in pages that are shared in
the first place, so that they do not need to be copied -- and moreover,
such objects will commonly never be modified any more.

@@

The type \lstinline!object_t!  is special as it causes the
compiler\footnote{Clang 3.5 with some patches to fix bugs in its address-space
 256 feature} to make all accesses through it relative to the $\%gs$
register.  With exceptions, nearly all accesses to objects managed by
the TM system should use this type so that the CPU will translate the
reference to the right version of the object.

@@

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction. \remi{so this is where we mention how the GC-STM
integration is useful. highlight more or move to own section?}

@@

@@

@@

The \textbf{slow path} is shown here:

@@

>>>>

\cfbolz{both the introduction and the contributions need to stress the co-design of GC and STM}

@@

When we need to store into our objects pointers to other objects, we
need to be careful: if we were to directly store regular addresses valid
in the local segment, then the objects' binary content would have to differ
between segments.  This would prevent any sharing.  Instead, we store
pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
of the segment.  A $SO$ can then be interpreted from any segment by
adding it to the base address of that segment.\footnote{An alternative
model would be to simulate threads using several independent processes
in which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the state
(file descriptors, external libraries, ...) is shared.} The result is
called a \emph{Linear Address (LA)}. This is illustrated in
figure~\ref{fig:Segment-Addressing}.

@@

Conflicts can be resolved by aborting one of the two transactions, or in
some cases by pausing one of them until the other commits.  If a commit
succeeds, we copy the changes to modified objects so that they become
visible in other segments.  Note that this only includes objects that
existed before the transaction and have been modified by it.  It does
not includes all the new objects created by the current transaction
(there is typically a very high rate of object churn in dynamic languages).
Our garbage collector allocates new objects in pages that are shared in
the first place, so that they do not need to be copied -- and moreover,
such objects will commonly never be modified any more.

@@

The type \lstinline!object_t!  is special as it causes the
compiler\footnote{We use Clang 3.5 with some patches to fix bugs in its address-space
 256 feature.} to make all accesses through it relative to the $\%gs$
register.  With exceptions, nearly all accesses to objects managed by
the TM system should use this type so that the CPU will translate the
reference to the right version of the object.

@@

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction. \remi{so this is where we mention how the GC-STM
integration is useful. highlight more or move to own section?}
\cfbolz{definitely its own section listing all the benefits}

@@

\cfbolz{maybe it's worth explicitly pointing out here that this is one of the
    few places where the \%gs translation must not be used}

@@

\cfbolz{is \%s translation used here? I guess yes, currently it's possible to misunderstand that}

@@

\cfbolz{put the slow path into a figure, if it straddles a page it's too hard figuring out indentation}
The \textbf{slow path} is shown here:

@@

\cfbolz{at this point I kind of had the need for a diagram of the various
states an object can be in, and the transitions between them.}
